---
title: "IBD / relatedness report"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    df_print: paged
params:
  genome_file: NULL
  pair_table: NULL
  sample_list: NULL
  exclude_list: NULL
  min_pi_hat: 0.1875
  flag_pi_hat: 0.35
  same_patient_pi_hat: 0.30
  duplicate_pi_hat: 0.95
  max_heatmap_samples: 100
  show_heatmap: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE,warning=FALSE,message=FALSE)
library(tidyverse)
library(ggplot2)
library(DT)
has_heatmaply<-requireNamespace("heatmaply",quietly=TRUE)

resolve_path<-function(param_path,default_name){
  if(!is.null(param_path)&&nzchar(param_path)&&file.exists(param_path))return(param_path)
  if(file.exists(default_name))return(default_name)
  NULL
}

pair_path<-resolve_path(params$pair_table,"pair.table")
sample_path<-resolve_path(params$sample_list,"sample.list")
exclude_path<-resolve_path(params$exclude_list,"exclude.list")

htmltools::tagList(htmltools::tags$style(htmltools::HTML(
".dataTables_wrapper table.dataTable td.dt-wrap{white-space:normal!important;word-break:break-word;font-size:0.88em;line-height:1.25;vertical-align:top}"
)))

ibd_dt<-function(x,pageLength=25){
  wrap<-intersect(c("category","action"),names(x))
  idx<-match(wrap,names(x))-1L
  defs<-if(length(idx))lapply(idx,function(i)list(targets=i,className="dt-wrap"))else list()
  datatable(x,rownames=FALSE,filter="top",
    options=list(pageLength=pageLength,scrollX=TRUE,autoWidth=TRUE,columnDefs=defs),
    class="compact stripe hover")
}
```

PLINK `--genome` output (`PI_HAT` = proportion IBD).

**Expected relatedness:** tumor vs its normal (from `pair.table` when present: two columns, **tumor** then **normal**, tab-separated; one normal may appear with many tumors). Without `pair.table`, same **patient prefix** (ID before the last `-`) is used as a fallback.

**Review:** high `PI_HAT` between samples that are **not** a listed tumor/normal pair and **not** the same prefix.

**Metrics exclusions:** if `exclude.list` exists (from `metrics.smk`), those samples are dropped from tables in this report only. The underlying PLINK / joint VCF run is unchanged.

```{r meta}
tn_pairs<-NULL
pair_roles<-tibble(sample=character(),role=character())

if(!is.null(pair_path)){
  raw<-read.table(pair_path,sep="\t",header=FALSE,stringsAsFactors=FALSE,comment.char="#")
  if(ncol(raw)<2)stop("pair.table needs at least two columns: tumor<TAB>normal")
  if(ncol(raw)==2)colnames(raw)<-c("tumor","normal")
  tn_pairs<-bind_rows(
    raw%>%transmute(IID1=as.character(tumor),IID2=as.character(normal)),
    raw%>%transmute(IID1=as.character(normal),IID2=as.character(tumor))
  )
  pair_roles<-bind_rows(
    raw%>%transmute(sample=as.character(tumor),role="tumor"),
    raw%>%transmute(sample=as.character(normal),role="normal")
  )%>%distinct(sample,.keep_all=TRUE)
}

cohort<-NULL
if(!is.null(sample_path)){
  cohort<-readLines(sample_path,warn=FALSE)
  cohort<-cohort[nzchar(trimws(cohort))]
}

excluded<-character()
if(!is.null(exclude_path)){
  excluded<-readLines(exclude_path,warn=FALSE)
  excluded<-trimws(excluded)
  excluded<-excluded[!grepl("^#",excluded) & nzchar(excluded)]
}
```

```{r load}
genome<-read.table(params$genome_file,header=TRUE,stringsAsFactors=FALSE,check.names=FALSE)
if(!all(c("IID1","IID2","PI_HAT")%in%names(genome))){
  stop("genome file must contain IID1, IID2, PI_HAT")
}

is_tn_pair<-function(i1,i2){
  if(is.null(tn_pairs))return(FALSE)
  any(tn_pairs$IID1==i1&tn_pairs$IID2==i2)
}

pairs<-genome %>%
  filter(!is.na(PI_HAT),is.finite(PI_HAT),PI_HAT>=0,IID1!=IID2) %>%
  {if(length(excluded))filter(.,!IID1%in%excluded,!IID2%in%excluded) else .} %>%
  mutate(
    patient1=sub("-[^-]+$","",IID1),
    patient2=sub("-[^-]+$","",IID2),
    same_patient=patient1==patient2,
    tumor_normal_pair=map2_lgl(IID1,IID2,is_tn_pair),
    PI_HAT=round(PI_HAT,4)
  )

related<-pairs %>%
  filter(PI_HAT>=params$min_pi_hat) %>%
  mutate(
    category=case_when(
      PI_HAT>=params$duplicate_pi_hat~"duplicate_or_near_identical",
      tumor_normal_pair~"expected_tumor_normal",
      same_patient&PI_HAT>=params$same_patient_pi_hat~"expected_same_patient_prefix",
      !same_patient&!tumor_normal_pair&PI_HAT>=params$flag_pi_hat~"flag_high_ibd_different_patient",
      !same_patient&!tumor_normal_pair~"review_moderate_ibd_different_patient",
      TRUE~"other"
    ),
    action=case_when(
      category=="duplicate_or_near_identical"~"Check dup/label",
      category=="expected_tumor_normal"~"Expected T/N",
      category=="expected_same_patient_prefix"~"Expected prefix",
      category=="flag_high_ibd_different_patient"~"Review swap/IBD",
      category=="review_moderate_ibd_different_patient"~"Review if odd",
      TRUE~"—"
    ),
    category_label=recode(category,
      duplicate_or_near_identical="Near duplicate",
      expected_tumor_normal="Tumor/normal",
      expected_same_patient_prefix="Same prefix",
      flag_high_ibd_different_patient="High IBD, diff pt",
      review_moderate_ibd_different_patient="Moderate IBD, diff pt",
      other="Other",
      .default=category
    )
  ) %>%
  left_join(pair_roles%>%rename(role1=role),by=c("IID1"="sample")) %>%
  left_join(pair_roles%>%rename(role2=role),by=c("IID2"="sample")) %>%
  arrange(desc(PI_HAT),category,IID1,IID2)

n_genome_samples<-length(unique(c(genome$IID1,genome$IID2)))
n_samples<-length(unique(c(pairs$IID1,pairs$IID2)))
has_pair_table<-!is.null(tn_pairs)
```

## Excluded samples (metrics QC)

```{r excluded-intro, echo=FALSE}
if(!is.null(exclude_path)){
  cat("Using ", exclude_path, ".\n\n", sep="")
}
```

Samples that failed metrics QC (`exclude.list`: median target &lt; 15×, &lt; 80% targets at 20×, PF HQ aligned reads &lt; 3M, or &gt; 10% targets at zero coverage). Omitted from pair tables in this report only.

```{r excluded}
if(length(excluded)==0){
  if(is.null(exclude_path)){
    cat("No exclude.list found; no metrics-based exclusions applied.\n")
  }else{
    cat("exclude.list is empty; all samples passed IBD QC thresholds.\n")
  }
}else{
  ex_tbl<-tibble(sample=sort(excluded))
  knitr::kable(ex_tbl,col.names="Excluded sample")
}
```

## Summary

```{r summary}
summary_tbl<-related %>%
  count(category_label,sort=TRUE,name="pairs") %>%
  rename(category=category_label)

meta_rows<-tibble(
  metric=c(
    "Samples in PLINK genome",
    "Excluded (metrics QC)",
    "Samples in this report",
    "Cohort sample.list loaded",
    "pair.table loaded",
    "Pairs (all non-self)",
    "Pairs reported (PI_HAT ≥ threshold)",
    "Flagged (different subject, high IBD)"
  ),
  value=c(
    as.character(n_genome_samples),
    ifelse(length(excluded)==0,"none",paste(sort(excluded),collapse=", ")),
    as.character(n_samples),
    ifelse(is.null(cohort),"no",as.character(length(cohort))),
    ifelse(has_pair_table,pair_path,"no (prefix fallback only)"),
    as.character(nrow(pairs)),
    as.character(nrow(related)),
    as.character(sum(related$category=="flag_high_ibd_different_patient"))
  )
)

knitr::kable(meta_rows,col.names=c("Metric","Value"))
knitr::kable(summary_tbl,col.names=c("Category","Pair count"))
```

## Per-sample: worst unrelated sharing

For each sample, **max PI_HAT** to another sample that is **not** a `pair.table` tumor/normal match and **not** the same prefix. Use this to spot a single worst partner per library.

```{r per-sample}
all_samples<-unique(c(pairs$IID1,pairs$IID2))
if(!is.null(cohort))all_samples<-union(all_samples,cohort)

unrelated_pairs<-pairs %>%
  filter(!tumor_normal_pair,!same_patient,PI_HAT>0)

per_sample<-bind_rows(
  unrelated_pairs%>%transmute(sample=IID1,other=IID2,PI_HAT),
  unrelated_pairs%>%transmute(sample=IID2,other=IID1,PI_HAT)
) %>%
  group_by(sample) %>%
  slice_max(PI_HAT,n=1,with_ties=FALSE) %>%
  ungroup() %>%
  left_join(pair_roles,by="sample") %>%
  rename(sample_role=role) %>%
  arrange(desc(PI_HAT))

if(nrow(per_sample)==0){
  cat("No unrelated pairs with PI_HAT > 0.\n")
}else{
  ibd_dt(per_sample,pageLength=25)
}
```

## Review: different subjects (high IBD)

```{r flags}
flags<-related %>% filter(category=="flag_high_ibd_different_patient")

if(nrow(flags)==0){
  cat("No pairs in this category.\n")
}else{
  ibd_dt(flags%>%select(IID1,IID2,role1,role2,patient1,patient2,PI_HAT,category=category_label,action))
}
```

## Expected tumor / normal (`pair.table`)

```{r tn}
expected_tn<-related %>% filter(category=="expected_tumor_normal")

if(!has_pair_table){
  cat("No pair.table supplied; section empty. Add pair.table (tumor<TAB>normal) when tumor/normal calling is in use.\n")
}else if(nrow(expected_tn)==0){
  cat("No tumor/normal pairs above threshold (check pair.table IDs match sample IDs in PLINK).\n")
}else{
  ibd_dt(expected_tn%>%select(IID1,IID2,role1,role2,PI_HAT,category=category_label,action))
}
```

## Expected: same prefix only (no `pair.table` row)

```{r prefix}
expected_px<-related %>% filter(category=="expected_same_patient_prefix")

if(nrow(expected_px)==0){
  cat("No same-prefix pairs above threshold.\n")
}else{
  ibd_dt(expected_px%>%select(IID1,IID2,patient1,patient2,PI_HAT,category=category_label,action))
}
```

## Near-duplicate

```{r duplicates}
dups<-related %>% filter(category=="duplicate_or_near_identical")

if(nrow(dups)==0){
  cat("No near-duplicate pairs.\n")
}else{
  ibd_dt(dups%>%select(IID1,IID2,patient1,patient2,PI_HAT,category=category_label,action))
}
```

## All reported pairs (filterable)

```{r all-pairs}
ibd_dt(
  related%>%select(IID1,IID2,role1,role2,patient1,patient2,TN=tumor_normal_pair,same_prefix=same_patient,PI_HAT,category=category_label,action),
  pageLength=50
)
```

## PI_HAT distribution

```{r dist}
pairs %>%
  filter(PI_HAT>0) %>%
  ggplot(aes(x=PI_HAT))+
  geom_histogram(bins=50,fill="steelblue",color="white")+
  geom_vline(xintercept=params$min_pi_hat,linetype=2,color="gray40")+
  geom_vline(xintercept=params$flag_pi_hat,linetype=2,color="firebrick")+
  theme_minimal(base_size=12)+
  labs(title="PI_HAT among sample pairs",x="PI_HAT",y="Count")
```

## Heatmap (subset)

```{r heatmap, eval=params$show_heatmap && has_heatmaply}
if(!params$show_heatmap){
  cat("Heatmap disabled (show_heatmap=FALSE).\n")
}else if(!has_heatmaply){
  cat("Package heatmaply not available.\n")
}else{
  library(heatmaply)
  heat_ids<-related %>%
    filter(category%in%c("flag_high_ibd_different_patient","duplicate_or_near_identical")) %>%
    arrange(desc(PI_HAT))
  if(nrow(heat_ids)==0)heat_ids<-related%>%arrange(desc(PI_HAT))%>%head(50)

  top_samples<-heat_ids %>%
    select(IID1,IID2) %>%
    pivot_longer(everything(),values_to="sample") %>%
    distinct(sample) %>%
    pull(sample)

  if(length(top_samples)>params$max_heatmap_samples){
    top_samples<-heat_ids %>%
      pivot_longer(c(IID1,IID2),values_to="sample") %>%
      group_by(sample) %>%
      summarise(max_pi=max(PI_HAT),.groups="drop") %>%
      arrange(desc(max_pi)) %>%
      head(params$max_heatmap_samples) %>%
      pull(sample)
  }

  ibd_sub<-pairs %>% filter(IID1%in%top_samples,IID2%in%top_samples)
  mat<-matrix(NA_real_,length(top_samples),length(top_samples),dimnames=list(top_samples,top_samples))
  for(i in seq_len(nrow(ibd_sub))){
    a<-ibd_sub$IID1[i]
    b<-ibd_sub$IID2[i]
    v<-ibd_sub$PI_HAT[i]
    mat[a,b]<-v
    mat[b,a]<-v
  }
  diag(mat)<-0

  heatmaply(
    mat,
    colors=colorRampPalette(c("white","#fee8c8","#e34a33"))(100),
    xlab="Sample",ylab="Sample",
    main=paste0("PI_HAT (",length(top_samples)," samples)"),
    show_dendrogram=FALSE,
    fontsize_row=max(6,min(10,220/length(top_samples))),
    fontsize_col=max(6,min(10,220/length(top_samples))),
    width=max(900,min(1400,40*length(top_samples))),
    height=max(700,min(1200,40*length(top_samples)))
  )
}
```

```{r heatmap-off, eval=params$show_heatmap && !has_heatmaply}
cat("Install heatmaply for subset heatmap, or set show_heatmap: false.\n")
```
