---
title: "MNP Results Report"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: united
    highlight: tango
params:
  csv_dir: "data/mnp"
  plan_dir: "data/mnp"
  summary_dir: "data/mnp"
  chroms: "1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,17,18,19,20,21,22"
---

**Run the Python script first** to build summary files (chr-by-chr, fast):

`python mnp_report_data.py -p data/mnp -c data/mnp -o data/mnp`

Then render this Rmd. Summary files: `mnp_report_missing_variants.txt`, `mnp_report_pair_lof.txt`, `mnp_report_summary_stats.txt`.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(tidyverse)
library(knitr)

# Columns we care about for display
display_cols <- c("ID", "Gene", "Variant.LoF_level", "Variant.Category", "Variant.Class",
  "Variant.Consequence", "HGVSc", "HGVSp", "AutoGVP",
  "Sample.Zyg", "Sample.Depth", "Sample.AltDepth", "Sample.AltFrac")
```

## Load summary files (from Python script)

Summary files are produced by `mnp_report_data.py` (chr-by-chr). Load them here.

```{r load-summary}
summary_dir <- params$summary_dir
missing_path <- file.path(summary_dir, "mnp_report_missing_variants.txt")
pair_path <- file.path(summary_dir, "mnp_report_pair_lof.txt")
stats_path <- file.path(summary_dir, "mnp_report_summary_stats.txt")

if (!file.exists(missing_path)) stop("Run: python mnp_report_data.py -p ", params$plan_dir, " -c ", params$csv_dir, " -o ", summary_dir)
if (!file.exists(pair_path)) stop("Missing ", pair_path)
if (!file.exists(stats_path)) stop("Missing ", stats_path)

missing_summary <- read_tsv(missing_path, show_col_types = FALSE)
pair_summary <- read_tsv(pair_path, show_col_types = FALSE)
stats <- read_tsv(stats_path, show_col_types = FALSE, col_names = c("name", "value"))
n_carrier_combos_total <- stats %>% filter(name == "n_carrier_combos_total") %>% pull(value) %>% as.integer()
n_carrier_combos_with_missing <- stats %>% filter(name == "n_carrier_combos_with_missing") %>% pull(value) %>% as.integer()
```

### Missing-variant summary

- Total (pair, carrier-sample) combinations expected: `r nrow(plan_carriers)`
- Combinations with at least one missing variant: `r nrow(missing_summary)`

```{r table-missing}
if (nrow(missing_summary) > 0) {
  kable(missing_summary, format = "html",
        caption = "Pairs and samples where one or more expected variant IDs are missing from the VEP CSV")
} else {
  "No missing variants: every carrier sample has id1, id2, and mnp_id rows in the VEP CSV."
}
```

## Per-pair LoF (from summary)

LoF comparison is precomputed by the Python script (first occurrence per variant per chrom).

### LoF summary

- Total MNP pairs (with at least one VEP row): `r nrow(pair_summary)`
- Pairs where MNP LoF = 1 but SNP(s) not: `r sum(pair_summary$mnp_pathogenic_snps_not == "True", na.rm = TRUE)`
- Pairs where SNP(s) LoF = 1 but MNP not: `r sum(pair_summary$snps_pathogenic_mnp_not == "True", na.rm = TRUE)`

### Pairs with LoF disagreement (MNP vs SNPs)

```{r table-disagree}
disagree <- pair_summary %>%
  filter(mnp_pathogenic_snps_not == "True" | snps_pathogenic_mnp_not == "True") %>%
  select(mnp_id, chrom, pos, id1, id2, id_mnp, lof_SNP1, lof_SNP2, lof_MNP,
         mnp_pathogenic_snps_not, snps_pathogenic_mnp_not)
if (nrow(disagree) > 0) {
  kable(disagree, format = "html", caption = "Pairs with LoF disagreement")
} else {
  "No pairs with LoF disagreement in this set."
}
```

## Key fields: variant table (first 100 rows from first available VEP CSV)

```{r table-key-fields}
vep_files <- list.files(params$csv_dir, pattern = "mnp_gt\\.vep\\.csv$", full.names = TRUE)
if (length(vep_files) == 0) {
  chroms <- as.character(strsplit(params$chroms, ",")[[1]])
  vep_files <- file.path(params$csv_dir, paste0("chr", chroms, ".mnp_gt.vep.csv"))
  vep_files <- vep_files[file.exists(vep_files)]
}
vep_sample <- if (length(vep_files) > 0) {
  read_csv(vep_files[1], col_types = cols(.default = col_character()), show_col_types = FALSE) %>% head(100)
} else { tibble() }
show_cols <- intersect(display_cols, names(vep_sample))
if (nrow(vep_sample) > 0 && length(show_cols) > 0) {
  kable(vep_sample %>% select(any_of(show_cols)), format = "html", caption = "Key fields (first 100 rows)")
} else {
  "No VEP CSV found or no display columns present."
}
```

